### 2011-06-04 docs/pubs/EIM-2011/abstract.tex
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

\begin{abstract}
  Effective discovery and integration of ecological data within data
  management systems requires rich semantic information that can
  describe and relate the types of information contained within
  disparate data sets. Within the Semtools project, we have developed
  approaches for expressing and representing semantic annotations of
  data sets for supplementing attribute and data-level metadata with
  terms drawn from domain-specific ontologies. Annotations provide a
  formal mechansism that can be used together with reasoning systems
  to enhance existing data discovery and integration approaches.  In
  this paper we describe extensions to the Ecological Metadata
  Language (EML) and associated tools for storing and using semantic
  annotations.  Specifically, we describe new user interface
  components implemented within the Morpho EML metadata editor for
  capturing user-supplied semantic annotations, extensions to the
  Metacata system for storing and accessing annotations and
  corresponding OWL-DL ontologies, and a new API within Metacat that
  uses annotation metadata to provide concept-based search and
  integration of data sets.
%   The Data Manager paper from the last EIM included semantic
%   foreshadowing in the 'future directions' section and this is a
%   logical continuation of that work. Previously, we saw that the DM
%   library enabled data integration tasks provided we knew exactly
%   which data attributes from exactly which datasets were "compatible"
%   with each other. Here "compatible" means not simply structurally
%   similar (e.g. decimal numbers) but also conceptually similar
%   (e.g. ocean water temperature in C). By including rich semantic
%   annotations we initially show simple integration capabilities where
%   the subject domain of our data corpus is relatively constrained
%   (i.e. SBC-LTER) and the observational models represented are
%   minimally variable (i.e. context excluded). Semantically-enahnced
%   query capabilities facilitate this 'smart integration' in multiple
%   ways.
%   First, the corpus is refined through compound concept-based
%   queries that exercise OBOE-compatible ontological subsumption
%   hierarchies to locate roughly equivalent attributes across
%   datasets. Additional query constraints are defined in relation to
%   the precise observational model formally expressed in OBOE. Finally,
%   the actual data values are interrogated so that only those that fall
%   in the desired range are returned. From this resultant subset of
%   disparate data objects we create a single synthetic data product
%   that represents a 'materialization' of the observations. Technical
%   discussion will include high-level descriptions of the
%   infrastructure and tools employed with focus on our development in
%   the rapidly evolving semantic frontier where scalability and best
%   practices are particularly germane and equally elusive.
\end{abstract}

>>>>

\begin{abstract}
  Effective discovery and integration of ecological data within data
  management systems requires rich semantic information that can
  describe and relate the types of information contained within
  disparate data sets. Within the Semtools project, we have developed
  approaches for expressing and representing semantic annotations of
  data sets for supplementing attribute and data-level metadata with
  terms drawn from domain-specific ontologies. Annotations provide a
  formal mechansism that can be used together with reasoning systems
  to enhance existing data discovery and integration approaches.  In
  this paper we describe extensions to the Ecological Metadata
  Language (EML) and associated tools for storing and using semantic
  annotations.  Specifically, we describe new user interface
  components implemented within the Morpho EML metadata editor for
  capturing user-supplied semantic annotations, extensions to the
  Metacat system for storing and accessing annotations and
  corresponding OWL-DL ontologies, and a new API within Metacat that
  uses annotation metadata to provide concept-based search and
  integration of data sets.
%   The Data Manager paper from the last EIM included semantic
%   foreshadowing in the 'future directions' section and this is a
%   logical continuation of that work. Previously, we saw that the DM
%   library enabled data integration tasks provided we knew exactly
%   which data attributes from exactly which datasets were "compatible"
%   with each other. Here "compatible" means not simply structurally
%   similar (e.g. decimal numbers) but also conceptually similar
%   (e.g. ocean water temperature in C). By including rich semantic
%   annotations we initially show simple integration capabilities where
%   the subject domain of our data corpus is relatively constrained
%   (i.e. SBC-LTER) and the observational models represented are
%   minimally variable (i.e. context excluded). Semantically-enahnced
%   query capabilities facilitate this 'smart integration' in multiple
%   ways.
%   First, the corpus is refined through compound concept-based
%   queries that exercise OBOE-compatible ontological subsumption
%   hierarchies to locate roughly equivalent attributes across
%   datasets. Additional query constraints are defined in relation to
%   the precise observational model formally expressed in OBOE. Finally,
%   the actual data values are interrogated so that only those that fall
%   in the desired range are returned. From this resultant subset of
%   disparate data objects we create a single synthetic data product
%   that represents a 'materialization' of the observations. Technical
%   discussion will include high-level descriptions of the
%   infrastructure and tools employed with focus on our development in
%   the rapidly evolving semantic frontier where scalability and best
%   practices are particularly germane and equally elusive.
\end{abstract}

### 2011-06-04 docs/pubs/EIM-2011/annotation.tex
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
\caption{Partial OBOE Annotation for Kelp sampling data. Shaded nodes represent ontological concepts; rectangular nodes are data table attibutes mapped to OBOE characteristics.}
\label{fig:kelp-mass-model}
\end{figure}

>>>>

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{images/oboe}
  \caption{The main classes and properties of the extensible
    observation ontology (OBOE). While shown here using UML, the model is
    defined using OWL-DL.}
  \label{fig:oboe}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
\caption{Partial OBOE annotation for Kelp sampling data. Shaded nodes represent ontological concepts; rectangular nodes are data table attibutes mapped to OBOE characteristics. Measurement standards (e.g., grams) and collection protocols are not illustrated.}
\label{fig:kelp-mass-model}
\end{figure}

### 2011-06-04 docs/pubs/EIM-2011/application.tex
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

\subsection{Concept query}
The semantic query interface primarily supports locating datasets by how well their observational models match the given criteria. Query criteria largely mirror the structure of an Annotation in that combinations of Entity, Characteristic and Protocol are specified and optionally compounded when increased precision is sought.

By leveraging the relationships defined and/or inferred from the
ontology we are able to increase recall beyond what is possible for
simple keyword-based searches
\cite{berkley09:_improv_data_discov_for_metad}. Broad queries return
direct matches and also n-depth subclass matches. The queries can be
quickly honed when using this browsing paradigm which allows rapid
exploration of the data repository without the oneous of defining
complete observational queries \emph{de novo}. Measurement templates
as defined in OBOE compatible ontologies enable a single concept to
proxy its constituent parts, namely the valid Characteristics of
certain Entities that can be measured with certain Protocols and
Standards. This short-hand query generation highlights one of the
compelling applications of using OBOE compatible ontologies and can
also be leveraged when authoring Annotations.  Using compound semantic
query criteria applies a fine-grained filter on the datasets that are
returned. Results may include only those datasets that include
measurements for a set of specific Characteristics of a particular
Entity. Furthermore, a query can specify that those measurements come
from precisely the \emph{same instance of that Entity} as belaboured
by the authors of the OBOE specification (Schildhauer here
somewhere??).

\subsection{Data query}
For even more precise recall, the OBOE model can be partially \emph{materialized} during the query stage after which a data range filter can be applied. Different techniques are available for merging the Annotation with the data that it describes, but for performance reasons a hybrid approach has been adopted in which premliminary search results from a concept-based query are collated and only that subset is materialized. The Data Manager Library is conveniently used to load raw data objects because our corpus is described using EML and the Annotation syntax - both of which inform the correct use of the Data Manager query features. For any data attributes that match the concept query criteria, we then verify that those attributes contain data values within the range specifed by the initial semantic data query (\figref{fig:metacat-query}).

\begin{figure*}[!t]
\centering
\includegraphics[width=1.0\textwidth]{images/metacat-query.png}
\caption{Semantic data query web interface. Data packages containing observations of Kelp Wet Mass less than or equal to 5 [grams] are returned. Additional search options and compund query criteria can be specified within the other tabs. Matches can be saved in the data cart for later exploration.}
\label{fig:metacat-query}
\end{figure*}

\subsection{Data integration}
The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration pursuits.
In addition to inspecting the data for values within a range and returning the packages that contain a match, the "smart" data integration feature goes further by constructing a synthetic data product that represents the complete results of the query both in terms of the attributes and the values that are returned. Each original data package match may have wildly different syntactic structures (e.g. column number, naming, order) but could share a subset of attributes that are asserted as semantically compatible. These compatible, in-common attributes then become the facets and metrics in the synthetic dataset with their values filtered accordingly (\figref{fig:integration}). While preliminary results have been quite promising - albeit academic - continued development of algorithms for determining compatibility of annotated measurements (e.g. gram and ounce are both mass units) and for mutating measurement values using ontologically-defined conversions (e.g. 1000 milligrams in a gram) will enable more thorough data integration.

>>>>

\begin{figure*}[!t]
\centering
\includegraphics[width=1.0\textwidth]{images/metacat-query.png}
\caption{Semantic data query web interface. Data packages containing observations of Kelp Wet Mass less than or equal to 5 [grams] are returned. Additional search options and compound query criteria can be specified within the other tabs. Matches can be saved in the data cart for later exploration.}
\label{fig:metacat-query}
\end{figure*}

\mypara{Concept query.} The semantic query interface primarily supports locating datasets by how well their observational models match the given criteria. Query criteria largely mirror the structure of an Annotation in that combinations of Entity, Characteristic and Protocol are specified and optionally compounded when increased precision is sought.

By leveraging the relationships defined and/or inferred from the
ontology we are able to increase recall beyond what is possible for
simple keyword-based searches
\cite{berkley09:_improv_data_discov_for_metad}. Broad queries return
direct matches and also n-depth subclass matches. The queries can be
quickly honed when using this browsing paradigm which allows rapid
exploration of the data repository without the oneous of defining
complete observational queries \emph{de novo}. Measurement ``templates''
as defined in OBOE compatible ontologies enable a single concept to
proxy its constituent parts, namely the characteristics of
particular entities that can be measured with a set of protocols and
standards. This short-hand query generation highlights a
compelling reason for using OBOE extension ontologies and can
also be leveraged when creating or editing semantic annotations in the Morpho interface.

Using compound semantic query criteria applies a finer-grained filter on the datasets that are
returned. Results can be restricted to only those datasets that include 
measurements for a set of specific characteristics of a particular
observational entity. Furthermore, a query can specify that those measurements come
from precisely the \emph{same instance of that entity}; a feature that fully exercises the comprensive observational structure expressed in the annotation.

\mypara{Data query.} For even more precise recall, the OBOE model can be partially \emph{materialized} during the query stage after which a data range filter can be applied. Different techniques are available for merging the annotation with the data that it describes, but for performance reasons a hybrid approach has been adopted in which premliminary search results from a concept query are collated and only that subset is materialized. Because our corpus is described using EML in conjunction with the annotation syntax, the Data Manager Library can handily load the described raw data while the annotation informs the correct use of the Data Manager query and filtering features. For any measurements that match the concept query criteria, we verify that those measurements (e.g., attributes) contain data values within the range specifed in the initial semantic data query and return the data packages that contain them (\figref{fig:metacat-query}).

\mypara{Data integration.} The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration pursuits. In addition to inspecting the data for values within a range and returning the datasets that contain a match, the ``smart'' data integration feature goes further by constructing a synthetic data product that represents the complete results of the query in terms of both the attributes and the values that are returned. Each original data object may have wildly different syntactic structures (e.g. column number, naming, order) but could share a subset of attributes that are semantically compatible as defined in accompanying annotations. These compatible, in-common attributes become the facets and metrics in a synthetic dataset; their values filtered accordingly (\figref{fig:integration}). While preliminary results have been quite promising - albeit academic - continued development of algorithms for determining compatibility of annotated measurements (e.g., gram and ounce are both mass units) and for mutating measurement values using ontologically-defined conversions (e.g., 1000 milligrams in a gram) will enable increasingly thorough, automated data integration practices.

### 2011-06-04 docs/pubs/EIM-2011/framework.tex
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

\section{Discovery and Integration Framework}
\label{sec:framework}

The Semtools project has focused development efforts on three main
components: a Java library for accessing and manipulating OBOE
ontology extensions and semantic annotations, an annotation plugin for
the Morpho metadata editor, and query extensions for the Metacat data
catalog. Below we briefly describe these components and give a brief
overview of the OBOE model (for more in-depth presentation of OBOE see
\cite{madin07:_ontol_for_descr_and_synth,bowers08}) and the semantic
annotation approach used in Semtools.

@@

\mypara{Semantic Annotations.} A semantic annotation consists of two
parts: (i) a ``configuration'' of the observation model containing the
specific entities, characteristics, observations, measurements, and so
on (drawn from one or more domain ontologies) that appropriately
capture the semantics of the data set; and (ii) a mapping between the
attributes in the data set to specific measurements defined in the
model configuration. \figref{fig:kelp-mass-model} shows a high-level
example of an annotation defined for a simple Kelp sampling data
set. Here, the data set consists of five attributes (bottom of
\figref{fig:kelp-mass-model}). Each attribute is mapped to a specific
measurement type (where only the characteristic of each measurement
type is shown), and measurement types are organized into observations
of specific Kelp entities (shown of type ``Macrocystis''), temporal
points (denoted by date-times), and spatial locations (given as site
names). Each measurement associated with a Kelp observation is assumed
to have occurred within the site and during the given time as
specified by the context relationships.  Semantic annotations provide
a formal description of attribute semantics, whereas in many
commonly-used metadata formats for describing data sets, largely only
informal text-based descriptions of data attributes are permitted.  In
the case of EML, there is some overlap between these two
mechanisms---particularly with respect to measurement
standards---however, semantic annotations extend this approach by
providing a general mechanism to formally associate concepts drawn
from domain ontologies to attributes. In the Semtools framework, we
employ an XML serialization syntax for semantic annotations that is
compatible with EML but that is stored separately from the EML
documentation of a data set (allowing, e.g., annotations to be used
independently of EML if needed). In addition, semantic annotations can
be used to ``materialize'' a given data set into a set of triples
conforming to the model configuration given in the annotation.
Materializing a data set to a set of OBOE triples in this way provides
a more uniform structural representation that can make a number of
integration tasks easier (described further below).

@@

\mypara{The Semantic Mediation API.} The Semantic Mediation API
includes basic ontology management features, annotation manipulation
capabilities, and simple concept navigation and visualization
components. The API is intended to be a centralized toolkit for use in
multiple application contexts (on either the client or server side).
The Semantic Mediation API uses both the OWL API \cite{} for ontology
management services including RDF/OWL parsing, serialization, and
simple class and property exploration as well as the Pellet
description-logic reasoner \cite{} for classification and exposing
inferred axioms in source ontologies. The inference services exposed
through the Semantic Mediation API is used in both discovery and
integration described below.
% is particularly powerful when utilizing OBOE Measurement 'templates'
% that express exactly what (Entity and Characteristic) can be
% observed and how (Standard and Protocol).
In our current Morpho and Metacat extensions, semantic annotations are
managed and stored automatically in an underlying, local relational
database. While it is also possible to use in-memory approaches for
storing and querying annotations, we found the overhead to be
prohibitive when large numbers of data sets are managed.

@@

\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
plugin for Morpho provides a front-end to the Semantic Mediation API
and allows users to define annotations for existing data package
descriptions. The editor provides a simple ``fill-in-the-blank'' style
form-based interface with a searchable hierachical concept selection
widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
integrates with a standard Morpho installation and provides semantic
query capabilities for locating data packages, marking up data
packages with semantic annotations, and saving annotations locally or
to a shared repository where they can be discovered and explored by
other users. The annotation editor in Morpho allows a user to view the
data set being annotated and fill in (by selecting an appropriate
ontology term) the characteristic, measurement standard, protocol, and
associated entity for each data-set attribute. Users can also specify
whether an observation spans multiple columns, and can provide context
relationships between attributes (i.e., obserations). The editor
provides a number of additional features including allowing users to
view the entire annotation (similar to \figref{fig:kelp-mass-model})
and to specify additional mapping constraints for observations and
measurements (not shown in \figref{fig:morpho-annotation}).

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{images/morpho-annotation-widget.png}
\caption{Morpho metadata editor with Semantic
  plugin. Fill-in-the-blank style editing interface to encourage
  natural language descriptions. A searchable hierarchical browser is
  used to select concepts from an ontology.}
\label{fig:morpho-annotation}
\end{figure*}

>>>>

\section{Semtools Framework}
\label{sec:framework}

The Semtools project has focused development efforts on three main
components: a Java library for accessing and manipulating OBOE
ontology extensions and semantic annotations, an annotation plugin for
the Morpho metadata editor, and query extensions for the Metacat data
catalog. Below we briefly describe these components and give a brief
overview of the OBOE model as well as the semantic
annotation approach used in Semtools. For a more in-depth presentation of OBOE see
\cite{madin07:_ontol_for_descr_and_synth,bowers08}.

@@

\mypara{Semantic Annotations.} A semantic annotation consists of two
parts: (i) a ``configuration'' of the observation model containing the
specific entities, characteristics, observations, measurements, and so
on (drawn from one or more domain ontologies) that appropriately
capture the semantics of the data set; and (ii) a mapping between the
attributes in the data set to specific measurements defined in the
model configuration. \figref{fig:kelp-mass-model} shows a high-level
example of an annotation defined for a simple Kelp sampling data
set. Here, the data set consists of five attributes (bottom of
\figref{fig:kelp-mass-model}). Each attribute is mapped to a specific
measurement type (where only the characteristic of each measurement
type is shown), and measurement types are organized into observations
of specific Kelp entities (shown of type ``Macrocystis''), temporal
points (denoted by date-times), and spatial locations (given as site
names). Each measurement associated with a Kelp observation is assumed
to have occurred within the site and during the given time as
specified by the context relationships.  Semantic annotations provide
a formal description of attribute semantics, whereas in many
commonly-used metadata formats for describing data sets,  only
informal text-based descriptions of data attributes are permitted.  In
the case of EML, there is some overlap between these two
mechanisms---particularly with respect to measurement
standards---however, semantic annotations extend this approach by
providing a general mechanism to formally associate concepts drawn
from domain ontologies to attributes. In the Semtools framework, we
employ an XML serialization syntax for semantic annotations that is
compatible with EML but that is stored separately from the EML
documentation of a data set (allowing, e.g., annotations to be used
independently of EML or with other metadata standards if needed). In addition, semantic annotations can
be used to ``materialize'' a given data set into a set of triples
conforming to the model configuration given in the annotation.
Materializing a data set to a set of OBOE triples in this way provides
a more uniform structural representation that can make a number of
integration tasks easier (described further below).

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=1.0\textwidth]{images/morpho-annotation-widget.png}
\caption{Morpho metadata editor with Semantic
  plugin. The fill-in-the-blank interface uses
  natural language descriptions for intuitive editing. A searchable, hierarchical browser is
  used to select concepts from domain-specific ontologies.}
\label{fig:morpho-annotation}
\end{figure*}

\mypara{The Semantic Mediation API.} The Semantic Mediation API
includes basic ontology management features, annotation manipulation
capabilities, and simple concept navigation and visualization
components. The API is intended to be a centralized toolkit for use in
multiple application contexts (on either client or server deployments).
The Semantic Mediation API uses both the OWL API \cite{owlapi} for ontology
management services including RDF/OWL parsing, serialization, and
simple class and property exploration as well as the Pellet
description-logic reasoner \cite{pellet} for classification and exposing
inferred axioms in source ontologies. The inference services exposed
through the Semantic Mediation API are used in both discovery and
integration described below.
% is particularly powerful when utilizing OBOE Measurement 'templates'
% that express exactly what (Entity and Characteristic) can be
% observed and how (Standard and Protocol).
In our current Morpho and Metacat extensions, semantic annotations are
managed and stored automatically in an underlying, local relational
database. While it is also possible to use in-memory approaches for
storing and querying annotations, we found the overhead to be
prohibitive when large numbers of data sets are managed.

@@

\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
plugin for Morpho provides a front-end to the Semantic Mediation API
and allows users to define annotations for existing data package
descriptions. The editor provides a simple ``fill-in-the-blank'' style
form-based interface with a searchable hierachical concept selection
widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
integrates with a standard Morpho installation and provides semantic
query capabilities for locating data packages, marking up data
packages with semantic annotations, and saving annotations locally or
to a shared repository where they can be discovered and explored by
other users. The annotation editor in Morpho allows a user to view the
data set being annotated and fill in (by selecting an appropriate
ontology term) the characteristic, measurement standard, protocol, and
associated entity for each data-set attribute. Users can also specify
whether an observation spans multiple columns, and can provide context
relationships between attributes (i.e., obserations). The editor
provides a number of additional features including allowing users to
view the entire annotation (similar to \figref{fig:kelp-mass-model})
and to specify additional mapping constraints for observations and
measurements.

### 2011-06-04 docs/pubs/EIM-2011/intro.tex
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

A number of recent efforts within the earth and environmental
informatics communities are adopting the notion of an observation as a
key modeling concept for enabling improved discovery and integration
of scientific data
\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,cushing07:_compon_based_end_user_datab,balhoff10:_phenex,bowers08}. These
approaches provide higher-level observational data models for
describing and representing observations and measurements found in
underlying datasets by defining common ``core'' concepts such as the
entities or features being observed, measurement units and protocols,
and context relationships between observations \cite{om,bowers08}.  A
major goal of these approaches is to enable interoperability and
uniform access to data by abstracting away the underlying
representation details that often exist across scientific data sets.

>>>>

A number of recent efforts within the earth and environmental
informatics communities are adopting the notion of an \emph{observation} as a
key modeling concept for enabling improved discovery and integration
of scientific data
\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,cushing07:_compon_based_end_user_datab,balhoff10:_phenex,bowers08}. These
approaches provide higher-level observational data models for
describing and representing observations and measurements found in
underlying datasets by defining common ``core'' concepts such as the
entities or features being observed, measurement units and protocols,
and context relationships between observations \cite{om,bowers08}.  A
major goal of these approaches is to enable interoperability and
uniform access to data by abstracting away the underlying
representation details that often exist across scientific data sets.

### 2011-06-04 docs/pubs/EIM-2011/summary.tex
resolve section references from intro.tex to other .tex files
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

\section{Related Work}

>>>>

\section{Summary}
\label{sec:summary}

The Semtools project has been successful in exploring and codifying technologies and techniques for applying semantic concepts to observational data. By providing mechanisms for linking datasets to ontological terms organized in a high-level observational model (e.g., OBOE), these new extensions to Metacat and Morpho help to overcome a number of limitations in
existing metadata management systems that strive to provide effective data discovery and integration features. Our close involvement with the SONet Project (Scientific Observations Network) \cite{sonet} encourages continued use-case refinement that will inform future semantic tool development and place an emphasis on intuitive interfaces and incremental adoption. This varied community of stakeholders is firmly invested in the use of cutting edge semantic solutions that will ultimately benefit multiple science disciplines by reducing obstacles to broad data sharing and innovative reuse.
